---
title: "Untitled"
author: "Jeffrey Arnold"
date: "5/31/2017"
output: html_document
---

```{r message=FALSE}
library("tidyverse")
library("stringr")
library("lubridate")
library("pollstR")
library("rstan")
```

All national polls in the Huffington Post Pollster database:
```{r}
polls <- pollster_charts_polls("2016-general-election-trump-vs-clinton") %>%
  `[[`("content") %>%
  # calculate the mid-date for date ranges of polls
  mutate(n_days = as.integer(difftime(end_date, start_date, units = "days")) + 1L,
         mid_date = start_date + ddays(ceiling(n_days / 2))) %>%
  # convert from 0-100 to 0-1
  mutate_at(vars(Trump, Clinton, Other, Undecided), funs(. / 100)) %>%
  # calculate shares that ignore missing values
  mutate(Clinton_Trump_share = Clinton + Trump,
         Clinton_adj = Clinton / Clinton_Trump_share,
         Trump_adj = Trump / Clinton_Trump_share) %>%
  # add polling standard errors
  mutate_at(vars(Clinton, Trump, Clinton_adj, Trump_adj), 
            funs(se = sqrt(. * (1 - .) / observations)))

```
The election result on November 8, 2016 (popular vote):
```{r}
election_result <- tibble(Clinton_n = 65853516,
                          Trump_n = 62984825,
                          total_votes = 136669237,
                          date = as.Date("2016-11-08")) %>%
  mutate(Clinton_adj = Clinton_n / (Clinton_n + Trump_n),
         Clinton = Clinton_n / total_votes)
                          
```

Let's plot the data:
```{r}
ggplot(filter(general_elec, mid_date > as.Date("2016-07-01")),
       aes(x = mid_date, y = Clinton_adj)) +
  geom_hline(yintercept = 0.5, colour = "white", size = 2) +
  geom_point() +
  labs(x = "", y = "Clinton vs. Trump")
```

A first cut. For each week, calculate mean and standard deviation of the pools.
```{r}
polls_weekly <-
  polls %>%
  mutate(week_date = ceiling_date(mid_date, "week") - ddays(3.5)) %>%
  filter(week_date > as.Date("2016-07-01"),
         week_date < as.Date("2016-11-08")) %>%
  group_by(week_date) %>%
  summarise(Clinton_adj = weighted.mean(Clinton_adj, observations),
            observations = sum(observations)) %>%
  mutate(Clinton_adj_se =
           sqrt(Clinton_adj * (1 - Clinton_adj) / observations))

ggplot() +
  geom_hline(yintercept = 0.5, colour = "white", size = 2) +  
  geom_pointrange(data = polls_weekly,
                  mapping = aes(x = as.Date(week_date),
                                y = Clinton_adj,
                                ymin = Clinton_adj - 2 * Clinton_adj_se,
                                ymax = Clinton_adj + 2 * Clinton_adj_se)) +
  geom_point(data = filter(polls, mid_date > as.Date("2016-07-01")), 
             mapping = aes(x = mid_date, y = Clinton_adj), 
             alpha = 0.5) +
  labs(x = "date", y = "Clinton vs. Trump")

```


```{r}
START_DATE <- as.Date("2016-07-01")
elec_jul_plus <- filter(general_elec, mid_date > START_DATE)
```
```{r}
theta1_prior <- 
  filter(general_elec, mid_date > as.Date("2016-06-15"),
         mid_date < as.Date("2016-07-01")) %>%
  summarise_at(vars(Clinton_adj), funs(mean, sd))

polling_data <- 
  within(list(), {
    y <- elec_jul_plus$Clinton_adj
    s <- elec_jul_plus$Clinton_adj_se
    N <- length(y)
    time <- as.integer(difftime(elec_jul_plus$mid_date, START_DATE),
                       units = "days") + 1L
    T <- as.integer(difftime(as.Date("2016-11-08"),
                             START_DATE, units = "days")) + 1L
    theta_init_loc <- theta1_prior$mean
    theta_init_scale <- theta1_prior$sd
  })
```

```{r}
mod_polling1 <- stan_model("polling1.stan")
```

```{r}
fit_polling1 <- sampling(mod_polling1, data = append(polling_data, list(tau = 0.02)),
         init = 0, chains = 1)
```

```{r}
time2date <- tibble(date = seq(START_DATE, election_result$date, "day")) %>%
  mutate(time = row_number())

theta_polling <-                     
  summary(fit_polling1, par = "theta")$summary %>%
  as.data.frame() %>%
  mutate(time = row_number()) %>%
  left_join(time2date, by = "time")

ggplot(theta_polling, aes(x = date, y = mean)) +
  geom_point() +
  labs(x = "", y = "Clinton v. Trump")

```


```{r}
mod_polling2 <- stan_model("polling2.stan")
```



Suppose we want to add house effects, for polling houses $h \in 1, \dots H$.
$$
\begin{aligned}[t]
y_i &\sim \mathsf{Normal}(\mu_i, s_i^2)  & i = 1, \dots, N \\
\mu_i &= \theta_{t[i]} + \eta_{h[i]} \\
\theta_1 &\sim \mathsf{Normal}(m_{\theta_1}, s_{\theta_1}) \\
\theta_t &\sim \mathsf{Normal}(\theta_{t - 1}, \tau^2) & t = 2, \dots, T \\
\eta_h &\sim \mathsf{Normal}(0, \zeta) & h = 1, \dots, H
\end{aligned}
$$
with priors on $\tau$ and $\zeta$,
$$
\begin{aligned}[t]
\tau &\sim \mathsf{HalfCauchy}(0, s_{\tau}) \\
\zeta &\sim \mathsf{HalfCauchy}(0, s_{\zeta})
\end{aligned}
$$

1. Interpret each of the following, and implement if possible.

    1. Model $y$ as:
    $$
    y_i \sim \mathsf{StudentT}(4, \mu_i, s_i) 
    $$
    2. Model $\theta_t$ as:
    $$
    \theta_t \sim \mathsf{StudentT}(4, \theta_{t - 1}, \tau) 
    $$
    3. Model $\tau$ as:
    $$
    \begin{aligned}[t]
    \theta_t &\sim \mathsf{Normal}(\theta_{t - 1}, \tau_t)  \\
    \log(\tau_t) &\sim \mathsf{Normal}(\gamma + \rho \tau_{t-1}, \sigma^2) & \rho \in     (0, 1)
    \end{aligned}
    $$

1. Currently house effects allow for a pollster to be biased. How would you model
    these generalizations?
    
    1. Party affiliation
    2. Some pollsters are have more variable polls than others
    3. Some pollsters use similar models and methods
    
2. We know the result of the actual election. Using this as a model of public opinion (voting intentions), how would you incorporate that into the model?

3. There are multiple sample subpopulations in the data. How would you model them?

4. We have treated the outcome as continuous. However, this is an approximation.

    1. How would you model the outcome as a proportion?
    2. There were other responses other than Clinton and Trump. In this data there
        is also "Other" and "Don't Know". How would you model these - or multiple candidates?
    
5. What if instead of this model, we estimated each week separately using `normal.stan`?

    1. Use the previous week's polls as a prior?
    


## State Level Analysis

If you wanted to do state level analysis ... here are the slugs of state-level charts for US election.
```{r}
us_elec_states <- 
  pollster_charts_iter(election_date = "2016-11-08",
                                .max_pages = 1000,
                                tags = "2016-president") %>%
  map_chr("slug") %>% 
  str_subset("trump-vs-clinton") %>% 
  `[`(!str_detect(., "2016-general-election"))
```

This would be necessary to model electoral votes and the election outcome.

